\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-09-27 chapters/covert_align.tex
package

Finally, we argue that the increased performance of WCBLE is partly due to the
fact that
tLDA is a suitable first-stage classifier.
Firstly, imposing a Toeplitz-covariance structure strongly regularizes the
problem~\cite{Sosulski2022,VanDenKerchove2022}, at the benefit of decoding performance.
Secondly, this method has a synergy with CBLE since both make the same
assumption about the short-time stationary of the EEG background noise
within an epoch.
CBLE does not retrain the first-stage classifier for each time shift, but
rather trains it once within the given window.
After training, the classifier parameters represent some information about the
expected ERP waveform and background noise.
By applying the trained classifier to different time shifts, it assumes this
ERP waveform can be shifted in time, but since the classifier's information about
the background noise was only obtained from the initial window, CBLE assumes
its properties do not vary throughout the epoch.
The block-Toeplitz covariance structure of tLDA also assumes that the
background noise represented by this covariance after subtracting the class
averages is stationary within the epoch~\cite{Sosulski2022}.

>>>>

%Finally, we argue that the increased performance of WCBLE is partly due to the
%fact that
%tLDA is a suitable first-stage classifier.
%Firstly, imposing a Toeplitz-covariance structure strongly regularizes the
%problem~\cite{Sosulski2022,VanDenKerchove2022}, at the benefit of decoding performance.
%Secondly, this method has a synergy with CBLE since both make the same
%assumption about the short-time stationary of the EEG background noise
%within an epoch.
%CBLE does not retrain the first-stage classifier for each time shift, but
%rather trains it once within the given window.
%After training, the classifier parameters represent some information about the
%expected ERP waveform and background noise.
%By applying the trained classifier to different time shifts, it assumes this
%ERP waveform can be shifted in time, but since the classifier's information about
%the background noise was only obtained from the initial window, CBLE assumes
%its properties do not vary throughout the epoch.
%The block-Toeplitz covariance structure of tLDA also assumes that the
%background noise represented by this covariance after subtracting the class
%averages is stationary within the epoch~\cite{Sosulski2022}.

### 2024-09-27 chapters/wcble.tex
package

\section{Results}

%Four settings are evaluated to assess both the impact of the iterative Woody
%procedure and the extraction of robust latency features.
%These are CBLE-median and WCBLE-median, where the median of the latency
%probability distribution is used as latency estimate, and CBLE-peak and
%WCBLE-peak, where the maximum of the score output is used as latency estimate.

\todo{
refer to later chapter for use on real erp data
}
\subsection{Data simulation}
To verify our approach, we simulated 16 channel \ac{eeg} data with standard
10-20 electrode positions using the \texttt{simulate\_evoked} method of the
MNE (version 1.8)~\cite{Gramfort2013} software package.
\Acp{erp} were generated by projecting a sine wave pulse source time course in a dipole in the
left hemisphere to the scalp electrodes using a boundary element method forward
model~\cite{Mosher1999} MNE's \texttt{fsaverage} source space and anatomy.
Pink temporal noise is generated by passing gaussian noise through an infinite
impulse response filter with transfer function
$\frac{B(z)}{A(z)} = \frac{1}{1 -1z^{-1}+0.15z^{-2}}$.
Pink spatial noise is also added, using a noise covariance matrix constructed from the
cosine distances between electrode positions and scaling it by the inverse of
the resulting spatial.
We refer to~\textcite{Gramfort2014} for implementation details.
%Evoked potentials are simulated for three different source time course waveforms,
%a gaussian, a morlet wavelet and a sinusoid pulse with variable latency,
%defined by the following formulas:
%\begin{equation}
%  f_\text{gaussian}(t,l) = a
%  \frac{1}{\varsigma\sqrt{2\pi}}e^{-\frac{1}{2}\frac{\left(t-l\right)^2}{\varsigma/2}}
%\end{equation}
%\begin{equation}
%  f_\text{morlet}(t,l) = ae^{-\left(\frac{t-l}{\frac{c}{2\pi f}}\right)^2}
%\end{equation}
The source time course is defined by the following function:
\begin{equation}
  s(t,l) =
  \begin{cases}
    a\sin\left(2\pi f\left(t-l\right)\right) & \text{if  $l-\frac{1}{2f} < t < l-\frac{1}{2f}$} \\
    0 & \text{otherwise}
  \end{cases}
\end{equation}
over time $t$, with latency $l$ and amplitude $a=\num{1e{-7}}$V and frequency
$f=4$.
%The standard deviation that determines the width gaussian $\varsigma=0.1s$.
%For the morlet wavelet, the number of cycles $c=3$ and the frequency $f=4Hz$.
%For the sine pulse, the frequency $f=4$

@@

%A sample of the simulated evoked data is displayed in Figures~\ref{fig:wcble/sim-gausian}, \ref{fig:wcble/sim-morlet} and
A sample of the simulated evoked data is displayed in Figures~\ref{fig:wcble/sim-sine}.
%\fullpagefig{figures/wcble/simulated-gaussian.pdf}{}{fig:wcble/sim-gaussian}
%\fullpagefig{figures/wcble/simulated-morlet.pdf}{}{fig:wcble/sim-morlet}
\fullpagefig{\includegraphics[width=\textwidth]{figures/wcble/simulated-sine.pdf}}{}{fig:wcble/sim-sine}
\todo{Properly align plots and label axes}

\subsection{Latency estimation}
\todo{refer to figure}
We compared our proposed \ac{wcble} algorithm to \ac{cble} on a latency
estimation task using the pearson correlation coefficient $\rho$\todo{equation} at different
jitter and \ac{snr} levels.
For both \ac{cble} and \ac{wcble}, target latencies where extracted in a
10-fold cross-validation scheme using the robust median latency method after
fitting the model.
\Ac{wcble}'s maximum iteration number was set to 64.
Results are presented in Figure~\ref{fig:wcble/results/latency}.

\begin{figure}
  \sansmath
  \input{figures/wcble/fig_latency.pgf} \\
  \input{figures/wcble/fig_accuracy.pgf}
  \caption{align results}
  \label{fig:wcble/results/latency}
\end{figure}
\todo{invert x axis}
\todo{caption}
\todo{properly label axes}
\todo{discuss results}

@@

\subsection{Classification}
\todo{refer to figure}
Similar to the previous experiment, classification accuracy between target and
non-target epochs is assessed.

\todo{insert results figure and discuss}

\section{Discussion}

Reason that accuracy latency estimation plateaus and accuracy drops for cble,
but that both plateau for wcble, is because of the alignment

\todo{mention cross validation}
\todo{remove lcmv}

\todo{In this work, we introduce a new latency method that leverages data-driven,
joint spatiotemporal filtering based on the LCMV beamformer~\cite{Vliet2016}. The algorithm
integrates the spatiotemporal filtering in Woody's template matching iteration
scheme and can easily be extended to multi-component settings.
Our method attempts to model the spatial and temporal activation of
each \ac{erp} component, as well as it's latency distribution, and takes into acount the
spatial and temporal correlation of the EEG background noise.}

\todo{with lda, it is in the single class case equal to shifting a spatiotemporal
lcmv beamformer over the epochs and can be used to extract latencies.
multiclass:}

\section{Conclusion}
Current work is focusing on the extension to multi-component algorithm, in a
similar fashion to ouyang

>>>>

\section{Results}
\label{sec:wcble/results}
\todo{refer to later chapter for results on real data}
\subsection{Data simulation}
To verify our approach, we simulated 16 channel \ac{eeg} data with standard
10-20 electrode positions using the \texttt{simulate\_evoked} method of the
MNE (version 1.8)~\cite{Gramfort2013} software package.
\Acp{erp} were generated by projecting a sine wave pulse source time course in a dipole in the
left hemisphere to the scalp electrodes using a boundary element method forward
model~\cite{Mosher1999} MNE's \texttt{fsaverage} source space and anatomy.
Pink temporal noise is generated by passing gaussian noise through an infinite
impulse response filter with transfer function
$\frac{B(z)}{A(z)} = \frac{1}{1 -1z^{-1}+0.15z^{-2}}$.
Pink spatial noise is also added, using a noise covariance matrix constructed from the
cosine distances between electrode positions and scaling it by the inverse of
the resulting spatial.
We refer to~\textcite{Gramfort2014} for implementation details.
%Evoked potentials are simulated for three different source time course waveforms,
%a gaussian, a morlet wavelet and a sinusoid pulse with variable latency,
%defined by the following formulas:
%\begin{equation}
%  f_\text{gaussian}(t,l) = a
%  \frac{1}{\varsigma\sqrt{2\pi}}e^{-\frac{1}{2}\frac{\left(t-l\right)^2}{\varsigma/2}}
%\end{equation}
%\begin{equation}
%  f_\text{morlet}(t,l) = ae^{-\left(\frac{t-l}{\frac{c}{2\pi f}}\right)^2}
%\end{equation}
The source time course is defined by the following function:
\begin{equation}
  s(t,l) =
  \begin{cases}
    a\sin\left(2\pi f\left(t-l\right)\right) & \text{if  $l-\frac{1}{2f} < t < l-\frac{1}{2f}$} \\
    0 & \text{otherwise}
  \end{cases}
\end{equation}
over time $t$, with latency $l$ and amplitude $a=\num{1e{-7}}$V and frequency
$f=4$.
%The standard deviation that determines the width gaussian $\varsigma=0.1s$.
%For the morlet wavelet, the number of cycles $c=3$ and the frequency $f=4Hz$.
%For the sine pulse, the frequency $f=4$

@@

%A sample of the simulated evoked data is displayed in Figures~\ref{fig:wcble/sim-gausian}, \ref{fig:wcble/sim-morlet} and
A sample of the simulated evoked data is displayed in Figures~\ref{fig:wcble/sim-sine}.
%\fullpagefig{figures/wcble/simulated-gaussian.pdf}{}{fig:wcble/sim-gaussian}
%\fullpagefig{figures/wcble/simulated-morlet.pdf}{}{fig:wcble/sim-morlet}
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/simulated-sine.pdf}
}{%
  Simulated target and non-target evoked data. Targets contain a sine wave
  pulse jittered by $\sigma$. 32 EEG channels with pink
  spatiotemporal noise were simulated using a forward head model, channel C3 is
  plotted here.
}{fig:wcble/sim-sine}
\todo{Properly align plots and label axes}

\subsection{Latency estimation}
\todo{refer to figure}
We compared our proposed \ac{wcble} algorithm to \ac{cble} on a latency
estimation task using the pearson correlation coefficient $\rho$\todo{equation} at different
jitter and \ac{snr} levels.
For both \ac{cble} and \ac{wcble}, target latencies where extracted in a
10-fold cross-validation scheme using the robust median latency method after
fitting the model.
\Ac{wcble}'s maximum iteration number was set to 64.
Results are presented in Figure~\ref{fig:wcble/results/latency}.
A 95\ confidence intervals on each measure was obtained through bootstrapping
with 1000 permutations.

\begin{figure}[!ht]
  \input{figures/wcble/fig_latency.pgf} \\
  \input{figures/wcble/fig_accuracy.pgf}
  \caption{align results}
  \label{fig:wcble/results/latency}
\end{figure}
\todo{properly label axes}
\todo{discuss results}

@@

\subsection{Classification}
\todo{refer to figure}
Analogous to the previous experiment, 10-fold cross-validated classification
accuracy between target and non-target epochs was assessed.
Here, \ac{wcble} was compared to \ac{cble}, but both also to their base classifier
\ac{tlda}.
\Ac{cble} and \ac{wcble} classifiers were implemented as described
in~\ref{sec:wcble/methods}, with both the wavelet transformed score time series
and square transformed latencies as input to a second-stage logistic regression
classifier with $L_2$-norm regularization ($C=0.2$).
\todo{ref figure}

The classification accuracy of both \ac{tlda} and \ac{cble} decreases with
$\$sigma$ and \ac{snr}, but \ac{cble} outperforms \ac{tlda}.
This difference is particularily large when $\sigma$ is higher ($\sigma=0.2$
and $\sigma=0.3$) and \ac{snr} is not yet high.
\Ac{wcble} always scored better than \ac{cble} and \ac{wcble}, except for
$\sigma=0.1$ and $\text{\ac{snr}} <27\text{db}$, at which point both \ac{cble}
and \ac{wcble} performed at chance level.
When considering the accuracy of \ac{wcble}, we see that it shows more of a
plateau as \ac{snr} increases.
\todo{insert results figure and discuss}

\section{Discussion \& conclusion}
In this work, we introduced a new \ac{erp} latency estimation and decoding
method that applies a spatiotemporal classification iteratively along the time
dimension of the data, refining the dataset at each iteration by aligning
trials.
The algorithm combines \acf{cble} with Woody's template matching iteration
scheme and is therefore named \acf{wcble}.
Our method attempts to model the spatial and temporal activation of
each \ac{erp} component, as well as it's latency distribution.
Through the firs-stage classifier, it takes into acount the
spatial and temporal structure of EEG background noise and the trials from the
classes to discriminate.
\section{Conclusion}

The key finding of the simulation study presented
in~\ref{sec:wcble/results} is that \ac{wcble} is robust to higher evoked
potential jitter and lower \ac{snr} than \ac{cble}.
It could therefore be a better candidate to apply in \ac{erp} analysis and
decoding in those cases where jittered \ac{erp} data is of interest.
Reason that accuracy latency estimation plateaus and accuracy drops for cble,
but that both plateau for wcble, is because of the alignment
As  the noise increases beyond a certain threshold, the improvement vanishes.
At this point, both methods perform poorly in both tasks, so if latencies
cannot be properly estimated, no increase in classification performance can be
gained by aligning in \ac{wcble}.

As shown, the proposed method is suitable for \ac{bci} decoding setting
as well as single-trial \ac{erp} latency analyses.
In the case where there is only one data class present, the first-stage
classifier can be replaced by a spatiotemporal filter.
Instead of outputting a classification score, this filter should now output a
metric representing the presence of a template response in the signal.
A suiting candidate is the spatiotemporal
\ac{lcmv}-beamformer~\cite{VanVliet2015}\todo{refer chapter}.
The activation pattern can be set to the average of the set of trials to
analyze, and it can be refined at each \ac{wcble} iteration by taking the
average of aligned trials.

Finally, we argue that \ac{tlda} is a suitable first-stage classifier.
Firstly, imposing a Toeplitz-covariance structure strongly regularizes the
problem~\cite{Sosulski2022,VanDenKerchove2022}, at the benefit of decoding performance.
Secondly, this method has a synergy with CBLE since both make the same
assumption about the short-time stationary of the EEG background noise
within an epoch.
CBLE does not retrain the first-stage classifier for each time shift, but
rather trains it once within the given window.
After training, the classifier parameters represent some information about the
expected ERP waveform and background noise.
By applying the trained classifier to different time shifts, it assumes this
ERP waveform can be shifted in time, but since the classifier's information about
the background noise was only obtained from the initial window, CBLE assumes
its properties do not vary throughout the epoch.
The block-Toeplitz covariance structure of \ac{tlda} also assumes that the
background noise represented by this covariance after subtracting the class
averages is stationary within the epoch~\cite{Sosulski2022}.

The main limitation of the proposed method is its supposed lack to handle
multi-component \ac{erp} data.
In real \ac{erp} analysis settings, multiple components are usually present,
and each component has a distinct contribution to class discriminatbility and
latency distribution.
A single-component method will yield issues in interpretability and convergence
in the presence of other components, as it can `lock-on' to a given component
cluster and flatten out the others.
This can be avoided, however, by the choice of a proper region of interest
isolating solely the component of interest.
Current work focuses on extending \ac{wcble} to a multi-component setting, on
the one hand to performance latency estimation and decoding performance on real
\ac{erp} data and on the other hand to apply it as an \ac{erp} component
separation method in the fashion of~\textcite{Oyang2017}.

The proposed implementation in theory also supports processing multi-class
data, but the convergence and correctness of the \ac{wcble} solution is yet to
be properly studied for this case.
